Double Standards in Judging Collective Action\thanks{}

### 2022-08-26 Documents/manuscript.tex
first complete draft

\date{July 15, 2022}

@@

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      \noindent Collective action is a powerful force driving social
change but often sparks contention about what actions are acceptable
means to effect social change. In five studies (total \emph{N} = 2,979),
we investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In three
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police and for or against
restricting abortion in the United States. We found evidence for
ideology-based, but not for identity-based, double standards:
Participants judged the same protest actions to be more acceptable when
the protesters' cause aligned with their own ideological position---but
showed no ingroup bias when judging collective action by ingroup and
outgroup members. Our findings have theoretical and practical
implications for understanding the often divided response to prominent
social movements.
    \end{abstract}
  \end{@twocolumnfalse}
  % \renewcommand{\abstractname}{Statement of Relevance}
  % \begin{@twocolumnfalse}
  %   \maketitle
  %   \begin{abstract}
  %   \noindent 
  %   \end{abstract}
  % \end{@twocolumnfalse}
]

@@

In this paper, we examine double standards in judging collective
action---that is, whether observers judge the same protest actions as
more or less acceptable depending on their own and the protesters' group
memberships (\emph{identity-based double standards}) or on the
protesters' cause and how it aligns with their own ideological positions
(\emph{ideology-based double standards}). In two studies, we first
develop an instrument of 25 controversial protest actions, based in item
response theory, to capture double standards in judging collective
action. In three preregistered experiments, we then use this instrument
to test for double standards in judging protest actions for workers'
rights in the United Kingdom and for or against defunding the police and
for or against restricting abortion in the United States. We find
evidence for ideology-based, but not for identity-based, double
standards: People judge the same protest actions to be more acceptable
when the protesters' cause aligns with their own ideological
positions---but do not show ingroup bias when judging collective action
by ingroup and outgroup members.

@@

In this research, item response theory serves two purposes. First, item
response theory provides a statistical framework for scale development.
In Studies 1 and 2, we select the most discriminating and informative
actions to build an instrument for capturing double standards in the
experiments. Second, item response theory provides a formal and
statistical definition of double standards in judging collective action
as respondents being more accepting of the same protest actions
depending on who the protesters and what they are protesting. In
Experiments 1 and 2, we operationalize double standards as the
differences in \(\theta_{(z)}\) between conditions which, as
\(\theta_{(z)}\) is \(z\)-standardized, correspond to Cohen's \(d\)
effect sizes.

@@

The first purpose of our research is to develop an instrument to measure
where people draw the line between acceptable and unacceptable forms of
collective action. In Study 1, we ask participants to generate more and
less extreme protest actions and compile a pool of protest actions for
scale development. In Study 2, we ask participants to rate how
acceptable they consider each of those protest action to be and use item
response theory to select the most discriminating and informative
protest actions for our instrument. In so doing, we build an instrument
that enables both present and future research to capture double
standards in judging collective action. In this way, our research
situates the distinction between normative and non-normative collective
action in the eye of the beholders and opens it to scientific
investigation.

The second purpose of our research is to test for double standards in
judging collective action---that is, whether people judge the same
protest actions as more acceptable if the protesters' are ingroup
members or the protesters' cause aligns with the ingroup's interests
(\emph{Hypothesis 1a}, \emph{1b}) or if the protesters' cause aligns
with their own system-justifying or system-challenging motives
(\emph{Hypotheses 2a}, \emph{2b}). In Experiment 1, we use the
instrument developed in Studies 1 and 2 to test for double standards in
how acceptable participants consider collective action for workers'
rights in the United Kingdom. By varying both the participants' and the
protesters' group memberships (working/middle class), Experiment 1
compares the relative evidence for identity-based (\emph{Hypothesis 1a},
\emph{1b}) and ideology-based double standards (\emph{Hypothesis 2a}).
Extending Experiment 1, Experiment 2 considers both system-challenging
and system-defending collective action in the context of the 2020 Black
Lives Matter protests in the United States. By varying the protesters'
cause (for/against defunding the police) as well as the participants'
and the protesters' group memberships (Black/White), Experiment 2
compares the relative evidence identity-based (\emph{Hypothesis 1a},
\emph{1b}) and ideology-based double standards (\emph{Hypothesis 2a},
\emph{2b}). Together, the preregistered experiments provide a
comprehensive test of the hypothesized double standards in judging
collective action.

@@

Expanding Experiment 1, Experiment 2 considered reactions to both
system-challenging and system-defending collective action and, in so
doing, provided a stronger test of the hypothesized ideology-based
double standards. Contrary to \emph{Hypothesis 2a} and the underlying
assumption of a universal system-justifying motive, we found that
participants considered the same protest actions to be \emph{more}
acceptable when judging system-challenging collective action (for
defunding the police) than when judging system-defending collective
action. This finding might reflect our liberal-leaning sample. In line
with \emph{Hypothesis 2b}, we found that participants who rejected
system-justifying beliefs considered the same protest action more
acceptable when judging system-challenging collective action (for
defunding the police). We did not, however, find evidence for
ideological symmetry in this relationship as participants who endorsed
system-justifying beliefs considered system-challenging and
system-defending collective action to be equally unacceptable. In
addition, we did not find evidence for the alternative, simpler
\emph{Hypothesis 3} that that participants would judge the same protest
actions to be acceptable when they supported the protesters' cause. In
this way, Experiment 2 provided evidence for ideology-based, but not for
identity-based, double standards in judging collective action.

@@

@@

We will measured moral concerns with the 36-item moral foundations
questionnaire (MFQ-2, Atari et al., 2022) which assesses to what extent
participants endorse concerns about care, equality, proportionality,
loyalty, authority, and purity (e.g., ``I believe chastity is an
important virtue.''; 1 = \emph{does not describe me at all}, 5 =
\emph{describes me extremely well}). We embedded three further attention
checks within the questionnaire (e.g., ``To show that you are paying
attention and giving your best effort, please select `moderately
describes me'.'').

@@

As in Experiments 1 and 2, we found no evidence for identity-based
double standards: Female participants did not consider protest actions
\emph{against} restricting abortion to be any more acceptable, on
average, than male participants or protest actions \emph{for}
restricting abortion (\(d = 0.14, [0.00, 0.29]\)). Instead, we again
found evidence for ideology-based double standards: Conservative
participants considered the same protest actions to be more acceptable,
on average, when protesters \emph{supported} restricting abortion
(\(d = 0.24, [0.07, 0.43]\)) while liberal participants considered the
same protest actions to be more acceptable when protesters
\emph{opposed} restricting abortion (\(d = 0.90, [0.72, 1.08]\)). This
pattern was consistent across subsamples as conservative women
(\(d = 0.30, [0.04, 0.55]\)) and, to a lesser extent, conservative men
(\(d = 0.19, [-0.07, 0.46]\)) considered the same protest actions to be
more acceptable when protesters \emph{supported} restricting abortion
and as both liberal men (\(d = 0.76, [0.49, 1.03]\)) and women
(\(d = 1.04, [0.78, 1.32]\)) considered the same protest actions to be
more acceptable when protesters \emph{opposed} restricting abortion.
This double standard was, however, more pronounced among liberal
participants (\(d = 0.66, [0.41, 0.92]\)). In this way, Experiment 3
provided evidence for both ideological symmetry and asymmetry in judging
collective action for and against restricting abortion.

@@

@@

Our research was motivated by real-world examples that seemed to suggest
that where people draw the line between acceptable and unacceptable
protest actions depends on \emph{who} the protesters are and \emph{what}
they are protesting---in other words, that there are double standards in
judging collective action. Our findings, however, showed that it is not
so simple. Contradicting our hypothesis of identity-based double
standards, participants in two preregistered experiments did not show
ingroup bias when judging controversial protest actions taken by ingroup
and outgroup members to advance their group's interests. Instead,
supporting our hypothesis of ideology-based double standards,
participants judged the same controversial protest actions to be more
acceptable when the protesters' cause aligned with their own
(system-challenging) ideological position. In the remainder of this
General Discussion, we consider some limiting conditions on our findings
and discuss theoretical and practical implications for understanding the
often divided response to collective action.

@@

Despite its strengths, our research is qualified by important
limitations. First, our research examined the hypothesized relationships
for three political causes---protecting workers' rights, defunding the
police, restricting abortion---and thus provided only limited evidence
that our findings generalize beyond those causes. Relying on few stimuli
to establish an effect threatens the validity and replicability of
research findings and constrains the generalizability of psychological
research (Yarkoni, 2022). Future research should sample a wider range of
causes to address this pervasive but often ignored problem (Judd et al.,
2012). This is particularly important when studying collective action as
recent theorizing (Jost et al., 2017) and research (Osborne et al.,
2019) highlighted the importance of differentiating between progressive
and conservative causes. Second, our research was based on samples from
two Western, Educated, Industrialized, Rich, and Democratic (WEIRD;
Henrich et al., 2010) contexts---the United Kingdom and the United
States---which limits the cross-cultural generalizability of our
findings.

@@

Most broadly, our results showed that, for many controversial protest
actions, there is no consensus on what is and is not an acceptable means
to advance a cause. Instead, this judgement seems to depend on what the
cause of a protest is and how it aligns with one's beliefs and values.
In this way, our results confirm our contention that the distinction
between normative and non-normative forms of collective action should be
situated in the eye of the beholder. By developing an instrument for
measuring where people draw the line between acceptable and unacceptable
forms of collective action, we provided a paradigm that we hope will
stimulate research on when and why people differ in making this
distinction.

After all, this distinction can be consequential. If people believe that
even extreme actions are acceptable means to advance a cause they
support, especially if they moralize their support for that cause
(Mooijman et al., 2018), they might support violent extremism. In this
way, our findings could explain why 45\% of Republicans supported the
storming of the U.S. Capitol on January 6, 2021 (YouGov, 2021) even
though it was illegal, destructive, and violent. If people believe that
even non-violent actions are unacceptable means to advance a cause they
oppose, they might support stifling dissent. In this way, our findings
could explain why Republicans responded to Black Lives Matter protests
by seeking to further criminalize disruptive protest (e.g., blocking
highways, Quinton, 2021) that might be most effective at challenging
injustice (Shuman et al., 2022, 2021). For those reasons, we encourage
researchers to use our paradigm to further investigate when and why
people apply different standards when judging collective action for
different causes.

@@

Opinion polls show deep divides not only about the ends of protests, but
also about the means of protest. For example, only 11\% of Republicans
and 29\% of White Americans---compared to 59\% of Democrats and 66\% of
Black Americans---considered kneeling during the national anthem---a
non-violent form of collective action---an appropriate means to protest
racist police violence (YouGov, 2017). Our research showed that, in
contrast to well-documented double standards in other domains, people do
not apply a different standard when judging collective action by ingroup
members or for a cause aligned with the ingroup's interests. Instead,
our research demonstrated that people consider the same protest actions
to be more acceptable when the cause of the protest aligns with their
own ideological positions---in other words, that there are partisan
double standards in judging collective action.

@@

@@

@@

@@

@@

@@

>>>>

@@

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      \noindent Collective action is a powerful force driving social
change but often sparks contention about what actions are acceptable
means to effect social change. In five studies (total \emph{N} = 2,979),
we investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In three
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police and for or against
restricting abortion in the United States. We found evidence for
ideology-based, but not for identity-based, double standards:
Participants judged the same protest actions to be more acceptable when
the protesters' cause aligned with their own ideological position---but
showed no consistent ingroup bias when judging collective action by
ingroup and outgroup members. Our findings have theoretical and
practical implications for understanding the often divided response to
prominent social movements.
    \end{abstract}
  \end{@twocolumnfalse}
  % \renewcommand{\abstractname}{Statement of Relevance}
  % \begin{@twocolumnfalse}
  %   \maketitle
  %   \begin{abstract}
  %   \noindent 
  %   \end{abstract}
  % \end{@twocolumnfalse}
]

@@

In this paper, we examine double standards in judging collective
action---that is, whether observers judge the same protest actions as
more or less acceptable depending on their own and the protesters' group
memberships (\emph{identity-based double standards}) or on the
protesters' cause and how it aligns with their own ideological positions
(\emph{ideology-based double standards}). In two studies, we first
develop an instrument of 25 controversial protest actions, based in item
response theory, to capture double standards in judging collective
action. In three preregistered experiments, we then use this instrument
to test for double standards in judging protest actions for workers'
rights in the United Kingdom and for or against defunding the police and
for or against restricting abortion in the United States. We find
evidence for ideology-based, but not for identity-based, double
standards: People judge the same protest actions to be more acceptable
when the protesters' cause aligns with their own ideological
positions---but do not show consistent ingroup bias when judging
collective action by ingroup and outgroup members.

@@

In this research, item response theory serves two purposes. First, item
response theory provides a statistical framework for scale development.
In Studies 1 and 2, we select the most discriminating and informative
actions to build an instrument for capturing double standards in the
experiments. Second, item response theory provides a formal and
statistical definition of double standards in judging collective action
as respondents being more accepting of the same protest actions
depending on who the protesters and what they are protesting. In
Experiments 1--3, we operationalize double standards as the differences
in \(\theta_{(z)}\) between conditions which, as \(\theta_{(z)}\) is
\(z\)-standardized, correspond to Cohen's \(d\) effect sizes.

@@

The first purpose of our research is to develop an instrument to measure
where people draw the line between acceptable and unacceptable forms of
collective action. In Study 1, we ask participants to generate more and
less extreme protest actions and compile a pool of protest actions for
scale development. In Study 2, we ask participants to rate how
acceptable they consider each of those protest action to be and use item
response theory to select the most discriminating and informative
protest actions for our instrument. In so doing, we build an instrument
that enables both present and future research to capture double
standards in judging collective action.

The second purpose of our research is to test for double standards in
judging collective action---that is, whether people judge the same
protest actions as more acceptable if the protesters' are ingroup
members or the protesters' cause aligns with the ingroup's interests
(\emph{Hypothesis 1a}, \emph{1b}) or if the protesters' cause aligns
with their own system-justifying or system-challenging motives
(\emph{Hypotheses 2a}, \emph{2b}). In Experiment 1, we use the
instrument developed in Studies 1 and 2 to test for double standards in
how acceptable participants consider collective action for workers'
rights in the United Kingdom. By varying both the participants' and the
protesters' group memberships (working/middle class), Experiment 1
compares the relative evidence for identity-based (\emph{Hypothesis 1a},
\emph{1b}) and ideology-based double standards (\emph{Hypothesis 2a}).
Extending Experiment 1, Experiment 2 considers both system-challenging
and system-defending collective action in the context of the 2020 Black
Lives Matter protests in the United States. By varying the protesters'
cause (for/against defunding the police) as well as the participants'
and the protesters' group memberships (Black/White), Experiment 2
compares the relative evidence for identity-based (\emph{Hypothesis 1a},
\emph{1b}) and ideology-based double standards (\emph{Hypothesis 2a},
\emph{2b}). Building on Experiments 1--2, Experiment 3 examines how
liberal and conservative men and women judge collective action for or
against restricting abortion in anticipation of the Supreme Court
overturning its \emph{Roe v. Wade} decision on reproductive rights in
the United States. By varying the protesters' cause as well as the
participants' ideology and gender, Experiment 3 compares the relative
evidence for identity-based (\emph{Hypothesis 1b}) and ideology-based
(\emph{Hypothesis 2b}) double standards in judging collective action
concerning a conservative cause in a sample balanced by gender and
ideology. Together, the preregistered experiments provide a
comprehensive test of the hypothesized double standards in judging
collective action.

@@

Expanding Experiment 1, Experiment 2 considered reactions to both
system-challenging and system-defending collective action and, in so
doing, provided a stronger test of the hypothesized ideology-based
double standards. Contrary to \emph{Hypothesis 2a} and the underlying
assumption of a universal system-justifying motive, we found that
participants considered the same protest actions to be \emph{more}
acceptable when judging system-challenging collective action (for
defunding the police) than when judging system-defending collective
action. This finding might reflect our liberal-leaning sample.

In line with \emph{Hypothesis 2b}, we found that participants who
rejected system-justifying beliefs considered the same protest action
more acceptable when judging system-challenging collective action (for
defunding the police). We did not, however, find evidence for
ideological symmetry in this relationship as participants who endorsed
system-justifying beliefs considered system-challenging and
system-defending collective action to be equally unacceptable. In
addition, we did not find evidence for the alternative, simpler
\emph{Hypothesis 3} that that participants would judge the same protest
actions to be acceptable when they supported the protesters' cause.

One reason why we did not find evidence for ideological symmetry might
be that participants judged protests for or against a progressive cause
and that \emph{opposing} a progressive cause is not equivalent to
\emph{supporting} a conservative cause. Another reason might be that our
sample skewed liberal and lacked committed conservatives who might
accept more extreme means to oppose a progressive cause. Experiment 3
addressed those limitations by testing for ideology-based double
standards in judging collective action for or against a conservative
cause in a balanced sample of conservatives and progressives.

@@

In Experiment 3, we investigated potential double standards in judging
collective action for or against restricting abortion in the United
States. That is, we tested whether liberal and conservative women and
men applied different standards when judging collective action opposing
or supporting further restricting or banning abortion. We conducted this
study in June 2022 when many expected the Supreme Court to overturn its
\emph{Roe v. Wade} decision that hitherto had guaranteed access to legal
abortions in the first 20--24 weeks of pregnancy in the United States.
We focused on restricting abortion as it is a conservative issue, as it
primarily affects women's reproductive rights, and as the anticipated
Supreme Court decision would prompt several states to restrict or ban
abortion.

By varying the participants' political orientation and the protesters'
cause, we tested the preregistered hypothesis that, in line with
\emph{symmetrical} ideology-based double standards, conservatives would
judge the same protest actions to be more acceptable if the protesters
supported restricting abortion and that liberals would judge the same
protest actions to be more acceptable if the protesters opposed
restricting abortion. We tested the alternative hypothesis that, in line
with \emph{asymmetrical} ideology-based double standards, conservatives
would judge all protest actions to be equally unacceptable while
liberals would judge the same protest actions to be more acceptable if
the protesters opposed restricting abortion. By varying the
participants' gender, we further tested the preregistered hypothesis
that, in line with identity-based double standards, female participants
would judge protest actions against restricting abortion to be more
acceptable, on average, than male participants or protest actions for
restricting abortion.

@@

We measured moral concerns with the 36-item moral foundations
questionnaire (MFQ-2, Atari et al., 2022) which assesses to what extent
participants endorse concerns about care, equality, proportionality,
loyalty, authority, and purity (e.g., ``I believe chastity is an
important virtue.''; 1 = \emph{does not describe me at all}, 5 =
\emph{describes me extremely well}). We embedded three further attention
checks within the questionnaire (e.g., ``To show that you are paying
attention and giving your best effort, please select `moderately
describes me'.'').

@@

As in Experiments 1 and 2, we found evidence for ideology-based double
standards: Conservative participants considered the same protest actions
to be more acceptable, on average, when protesters \emph{supported}
restricting abortion (\(d = 0.24, [0.07, 0.43]\)) while liberal
participants considered the same protest actions to be more acceptable
when protesters \emph{opposed} restricting abortion
(\(d = 0.90, [0.72, 1.08]\)). This double standard was more pronounced
among liberal participants (\(d = 0.66, [0.41, 0.92]\)). In this way,
Experiment 3 provided evidence for both ideological symmetry and
asymmetry in judging collective action for and against restricting
abortion.

We found some evidence for identity-based double standards: Female
participants considered protest actions \emph{against} restricting
abortion to be more acceptable, on average, than male participants or
protest actions \emph{for} restricting abortion
(\(d = 0.14, [0.00, 0.29]\)). This pattern was, however, overshadowed by
stronger ideology-based double standards that were consistent across
subsamples: Conservative women (\(d = 0.30, [0.04, 0.55]\)) and, to a
lesser extent, conservative men (\(d = 0.19, [-0.07, 0.46]\)) considered
the same protest actions to be more acceptable when protesters
\emph{supported} restricting abortion and both liberal women
(\(d = 1.04, [0.78, 1.32]\)) and men (\(d = 0.76, [0.49, 1.03]\))
considered the same protest actions to be more acceptable when
protesters \emph{opposed} restricting abortion. In this way, Experiment
3 provided more consistent evidence for ideology-based than for
identity-based double standards in judging collective action.

@@

We explored whether group identification moderated participants'
judgements. We found, however, that gender identification neither
affected how women judged protesters opposing
(\(\beta_{xy} = -0.04, [-0.14, 0.07]\)) or supporting
(\(\beta_{xy} = -0.06, [-0.17, 0.03]\)) abortion restrictions nor how
men judged protesters opposing (\(\beta_{xy} = -0.05, [-0.14, 0.04]\))
or supporting (\(\beta_{xy} = -0.04, [-0.12, 0.06]\)) abortion
restrictions. Our non-preregistered analyses thus suggested that gender
identification did not moderate identity-based double standards in
judging collective action concerning reproductive rights.

@@

Our research was motivated by real-world examples that seemed to suggest
that where people draw the line between acceptable and unacceptable
protest actions depends on \emph{who} the protesters are and \emph{what}
they are protesting---in other words, that there are double standards in
judging collective action. Our findings, however, showed that it is not
so simple. Contradicting our hypothesis of identity-based double
standards, participants in three preregistered experiments did not show
consistent ingroup bias---in terms of class, race, and gender---when
judging controversial protest actions taken by ingroup and outgroup
members to advance their group's interests. Instead, supporting our
hypothesis of ideology-based double standards, participants judged the
same controversial protest actions to be more acceptable when the
protesters' cause---protecting workers' rights, defunding the police,
restricting abortion---aligned with their own ideological position. In
the remainder of this General Discussion, we consider some limiting
conditions on our findings and discuss theoretical and practical
implications for understanding the often divided response to collective
action.

@@

Despite its strengths, our research is qualified by several limitations.
First, our research examined the hypothesized relationships for three
political causes---protecting workers' rights, defunding the police,
restricting abortion---and thus provided only limited evidence that our
findings generalize beyond those causes. Relying on few stimuli to
establish an effect threatens the validity and replicability of research
findings and constrains the generalizability of psychological research
(Yarkoni, 2022). Future research should sample a wider range of causes
to address this pervasive but often ignored problem (Judd et al., 2012).
This is particularly important when studying collective action as recent
theorizing (Jost et al., 2017) and research (Osborne et al., 2019)
highlighted the importance of differentiating between progressive and
conservative causes. Second, our research was based on samples from two
Western, Educated, Industrialized, Rich, and Democratic (WEIRD, Henrich
et al., 2010) contexts---the United Kingdom and the United
States---which limits the cross-cultural generalizability of our
findings.

@@

Growing evidence shows that, in many situations, observers apply a more
lenient standards when judging moral transgressions by ingroup members
(Abrams et al., 2013; Endevelt et al., 2021). And yet, our findings
suggest that, unlike judgements in other domains, judgements about
collective action are not subject to double standards based on group
identities such as class, race, or gender---not even among highly
identified group members. One reason might be that collective action,
even in its controversial forms, is not judged as a moral transgression.
Instead, observers might understand collective action primarily as a
means to an end and judge how acceptable a form of collective action is
in relation to its end. This aligns with our finding that how acceptable
participants judged protest actions to be depended on the protesters'
cause, not their group membership.

Our findings showed that progressives and, to a lesser extent,
conservatives consider the same controversial protest actions to be more
acceptable when the protesters' cause aligns with their ideological
position. So far, we explained this ideology-based double standard as
resulting from conservatives being motivated to defend the system,
progressives being motivated to challenge the system, and both being
motivated to support collective action to those effects (Jost et al.,
2017; Osborne et al., 2019). An alternative explanation is to understand
those double standards as an expression of ingroup bias based on an
opinion-based (McGarty et al., 2009) or partisan (Finkel et al., 2020)
identity. This aligns with Verkuyten et al.'s (2022) finding that
participants were more tolerant of the same transgressive protest
actions when taken by their most-liked, rather than their least-liked,
political group (e.g., climate activists). Another explanation is that
their different moral concerns (Graham et al., 2009; Kivikangas et al.,
2021) lead progressives and conservatives to see their support of
certain causes as a fundamental matter of right or wrong (Skitka et al.,
2021) and, by thus moralizing those causes, to accept more extreme means
to achieving them. This aligns with Richardson \& Conway's (2022)
finding that their different moral concerns explained why liberals rated
protesting for liberal causes as more moral than conservatives and vice
versa. Future research should seek to establish which of those plausible
mechansims best explains partisan differences in judging collective
action.

Our findings raised the question why double standards in judging
collective action were more pronounced among liberal than among
conservative participants. One reason for this ideological asymmetry
might be that collective action that violates prevailing norms or laws
is seen as inherently threatening to the status quo, even if its aim is
to bolster and defend the system. Political conservatism can be
understood as having two psychologically related but distinguishable
dimensions, resistance to change and justification of inequality (Jost
et al., 2003). In this way, conservatives might be expected to condemn
disruptive protest even if the protest's cause is to defend inequality.
Future research should seek to establish how consistent this ideological
asymmetry is across different political causes and cultural contexts.

Most broadly, our findings showed that, for many controversial protest
actions, there is no consensus on what is and is not an acceptable means
to advance a cause. Instead, this judgement seems to depend on what the
cause of a protest is and how it aligns with one's beliefs and values.
In this way, our results confirm our contention that the distinction
between normative and non-normative forms of collective action should be
situated in the eye of the beholder. By developing an instrument for
measuring where people draw the line between acceptable and unacceptable
forms of collective action, we provide a paradigm that we hope will
stimulate research on when and why people differ in making this
distinction.

After all, this distinction is often consequential. If people believe
that even extreme actions are acceptable means to advance a cause they
support, especially if they moralize their support for that cause
(Mooijman et al., 2018), they might support violent extremism. In this
way, our findings could explain why 45\% of Republicans supported the
storming of the U.S. Capitol on January 6, 2021 (YouGov, 2021) even
though it was illegal, destructive, and violent. If people believe that
even non-violent actions are unacceptable means to advance a cause they
oppose, they might support stifling dissent. In this way, our findings
could explain why Republicans responded to Black Lives Matter protests
by seeking to further criminalize disruptive protest (e.g., blocking
highways, Quinton, 2021) that might be most effective at challenging
injustice (Shuman et al., 2022, 2021). For those reasons, we encourage
researchers to use our paradigm to further investigate when and why
people apply different standards when judging collective action for
different causes.

@@

Opinion polls show deep divides not only about the ends of protests, but
also about the means of protest. For example, only 11\% of Republicans
and 29\% of White Americans---compared to 59\% of Democrats and 66\% of
Black Americans---considered kneeling during the national anthem---a
non-violent form of collective action---an appropriate means to protest
racist police violence (YouGov, 2017). Our research showed that, in
contrast to well-documented double standards in other domains, people do
not apply a different standard when judging collective action by members
of the same class, race or gender or for a cause aligned with their own
group's interests. Instead, our research demonstrated that people
consider the same protest actions to be more acceptable when the cause
of the protest aligns with their own ideological positions---in other
words, that there are partisan double standards in judging collective
action.

@@

\leavevmode\vadjust pre{\hypertarget{ref-finkel_political_2020}{}}%
Finkel, E. J., Bail, C. A., Cikara, M., Ditto, P. H., Iyengar, S., Klar,
S., Mason, L., McGrath, M. C., Nyhan, B., Rand, D. G., Skitka, L. J.,
Tucker, J. A., Van Bavel, J. J., Wang, C. S., \& Druckman, J. N. (2020).
Political sectarianism in {America}. \emph{Science}, \emph{370}(6516),
533--536. \url{https://doi.org/10.1126/science.abe1715}

@@

\leavevmode\vadjust pre{\hypertarget{ref-graham_liberals_2009}{}}%
Graham, J., Haidt, J., \& Nosek, B. A. (2009). Liberals and
conservatives rely on different sets of moral foundations. \emph{Journal
of Personality and Social Psychology}, \emph{96}(5), 1029--1046.
\url{https://doi.org/10.1037/a0015141}

\leavevmode\vadjust pre{\hypertarget{ref-henrich_weirdest_2010}{}}%
Henrich, J., Heine, S. J., \& Norenzayan, A. (2010). The weirdest people
in the world? \emph{Behavioral and Brain Sciences}, \emph{33}(2-3),
61--135. \url{https://doi.org/10.1017/S0140525X10000725}

@@

\leavevmode\vadjust pre{\hypertarget{ref-kivikangas_moral_2021}{}}%
Kivikangas, J. M., Fernández-Castilla, B., Järvelä, S., Ravaja, N., \&
Lönnqvist, J.-E. (2021). Moral foundations and political orientation:
{Systematic} review and meta-analysis. \emph{Psychological Bulletin},
\emph{147}(1), 55--94. \url{https://doi.org/10.1037/bul0000308}

@@

\leavevmode\vadjust pre{\hypertarget{ref-mcgarty_collective_2009}{}}%
McGarty, C., Bliuc, A.-M., Thomas, E. F., \& Bongiorno, R. (2009).
Collective action as the material expression of opinion-based group
membership. \emph{Journal of Social Issues}, \emph{65}(4), 839--857.
\url{https://doi.org/10.1111/j.1540-4560.2009.01627.x}

@@

\leavevmode\vadjust pre{\hypertarget{ref-richardson_standing_2022}{}}%
Richardson, I., \& Conway, P. (2022). Standing up or giving up? {Moral}
foundations mediate political differences in evaluations of {BLACK}
{LIVES} {MATTER} and other protests. \emph{European Journal of Social
Psychology}, \emph{52}(3), 553--569.
\url{https://doi.org/10.1002/ejsp.2837}

@@

\leavevmode\vadjust pre{\hypertarget{ref-skitka_psychology_2021}{}}%
Skitka, L. J., Hanson, B. E., Morgan, G. S., \& Wisneski, D. C. (2021).
The psychology of moral conviction. \emph{Annual Review of Psychology},
\emph{72}(1), 347--366.
\url{https://doi.org/10.1146/annurev-psych-063020-030612}
